Validation of Visual Statistical Inference, Applied to Linear Models

### 2013-02-19 paper_jasa.tex
reduce pages by restructuring figures and moving proof of lemma to supplimentary

\begin {abstract}  
Statistical graphics play a crucial role in exploratory data analysis, model checking and diagnosis. A new method, called the lineup protocol, enables determining the statistical significance of visual findings, which bridges a big gulf between exploratory and inferential statistics.  In this paper inferential methods for statistical graphics is developed further by refining the terminology of visual inference, and framing the lineup protocol in a context that allows direct comparison with conventional tests, in scenarios when a conventional test exists. This framework is used to compare the performance of the lineup protocol against conventional statistical testing in the scenario of fitting linear models. A human subjects experiment is conducted using simulated data to provide controlled conditions. Results suggest that the lineup protocol provides comparable results to the conventional tests, and expectedly out-performs them when data is contaminated, a scenario where assumptions required for performing a conventional are violated, and, there may be some super-individuals who yield better power than the conventional test even in the most difficult of tasks.

@@

\begin{proof}
%We will now make use of properties of the data going into the lineup and assume that the properties are reflected in the lineup display:

 By definition

$$p_D=Pr\left(|t| \ge t_{obs} \mid H_0\right)=1-F_{|t|}(t_{obs}) \ \ \Rightarrow \ \  |t_{obs}|=F_{|t|}^{-1}(1-p_D)$$

%Now suppose $F_{|t|}(.;\delta)$ denotes the distribution function of an absolute value of $t$, the conventional test statistic.
%, with non-centrality parameter $\delta$. 
\noindent Then the distribution function of the $p$-value, $p_D$, under $H_0$, is uniform, since:  
\begin{eqnarray}\label{dist_p}
F_{p_D}(p) &=& Pr(p_D \le p)=1-Pr(1-p_D \le 1-p) \nonumber \\
  &=& 1-Pr\left(F_{|t|}^{-1}(1-p_D) \le F_{|t|}^{-1}(1-p) \right) \nonumber \\
  &=& 1-Pr\left(|t_{obs}| \le F_{|t|}^{-1}(1-p)\right) \nonumber \\
  &=&%\left\{ \begin{array}{ll}
          1-F_{|t|}\left( F_{|t|}^{-1}(1-p)\right)=p \mbox{ ; under $H_0$} 
 %         1-F_{|t|}( F_{|t|}^{-1}(1-p); \delta) &\mbox{ ; under $H_1$} 
%       \end{array} \right.     
\end{eqnarray}

%Thus the density of $p_D$ is Uniform(0,1)  under $H_0$. As noted by \cite{Ruppert:2007}, under $H_1$ the density of $p_D$ $$f_{p_D}(p_D; \delta)= \frac{f_{|t|}(F_{|t|}^{-1}(1-p_D);\delta)}{f_{|t|}(F_{|t|}^{-1}(1-p_D))}$$ derived from equation \eqref{dist_p}  is a right skewed distribution.

Let $p_{0,i}$, $i=1, ..., m-1$ denote the  $p$-values associated with data corresponding to the $m-1$ null plots. Since this data is generated consistently with the null hypothesis,  the $p$-values are independent and  follow a standard Uniform distribution, $p_{i,0} \sim U[0,1], i= 1, ..., m-1$. The minimum $p_0 = \min_{1 \le i \le m-1}  \ p_{0,i}$ then follows a Beta distribution with shape parameters 1 and $m-1$, and corresponding distribution function 
\[
F_{p_0} (x) =  1- (1- x) ^{m-1}  \text{ for } x \in [0,1].
\]

\noindent Thus
\begin{eqnarray*}
P(p_D < p_{0}) &=& 1 - P(p_{0} \le p_D) = 1- \int_0^1  P(p_{0} \le p_D \mid p_D=t) f_{p_D}(t) dt   \\
&=& 1 - \int_0^1 F_{p_{0}}(t) f_{p_D}(t) dt = 1 - \int_0^1 f_{p_D}(t) dt + \int_0^1 (1-t)^{m-1} f_{p_D}(t) dt  \\
&=&  E\left[ (1 - p_D)^{m-1}\right].
\end{eqnarray*}

%\noindent This proves the statement above. 
%We further see:
%\[
%E\left[ (1 - p_D)^{m-1}\right] =  \sum_{k=0}^{m-1} {m-1 \choose k} (-1)^k E[p_D^k] = 1-(m-1)E[p_D] + O(E[p_D^2]).
%\]

% If the density of $p$-value is very right skewed, the expectation term would be large. The distribution function would be highly right skewed when there would be strong signal in the data plot.Thats where the distribution function of $p$-value under alternative comes in. Should we add that in Equation \ref{dist_p}? )},
%conversely, as $m$, the number of choices given in the lineup, goes up the probability to pick the data plot  goes down. 
%}
%\blue{This shows that the lineup size $m$ has an effect on the power of the visual test: it is the $m-1$th moment of the distribution of $1-p_D$.

%
%
%
%Let us think of a lineup as a head-to-head comparison of the test statistic and $m-1$ null plots. 
%%Let the data plot have a $p$-value of $p_D$, and the null plots $p$ values of $p_{0, i}$ with $1 \le i \le m-1$.
%We know that for each comparison the probability that the data, on which the plot is based, has the smaller $p$ value is 
%\begin{eqnarray*}
%P(p_D < p_{0,i}) &=& 1 - P(p_{i,0} \le p_D) = 1- \int_0^1  P(p_{i,0} \le p_D \mid p_D=t) f_{p_D}(t) dt =  \\
%&=& 1 - \int_0^1 F_{p_{0,i}}(t) f_{p_D}(t) dt = 1 - \int_0^1 t f_{p_D}(t) dt = 1 - E[p_D],
%\end{eqnarray*}
%which, in particular, is independent of $p_{0,i}$ for all $i$.
%
%Let us make the assumption that an observer is able to identify  the chart corresponding to the data with the smallest $p$-value. Further we will assume that all observers  have the same ability in identifying the data plot.
%
%With that, we define $Z$  as the number of null plots in a lineup, for which the $p$-value $p_{0,i}$ is smaller than $p_D$.  
%
%Then $Z \sim B_{m-1, E[p_D]}$, and the probability that an observer will pick the data plot in a given lineup is 
%\[
%P(Z=0) = \left(1 - E[p_D] \right)^{m-1} = P(p_D \le p_0), \ \ \ \text{ where } p_0 = \min_{1 \le i \le m-1}  \ p_{0,i}.
%\]
\end{proof}

@@

Figures \ref{fig:P-val_log} and \ref{fig:P-val_log2} show the selection counts and $p$-values of all plots for each lineup in experiments 1 and 2, respectively. In each of these plots, the log of $p$-value for each plot in the lineup is plotted horizontally and the count of the number of subjects who selected this plot is displayed vertically. The dot represents this point, and the lines connect vertically with 0 count. Columns are ordered according to effect size as defined in section 6.2; rows show replicates for the same combination of parameters.

@@

\begin{figure*}[hbtp]
   \centering
      \includegraphics[width=0.95\textwidth]{p_val_log_counts-a.pdf}
       \includegraphics[width=0.95\textwidth]{p_val_log_counts-b.pdf}
       \caption{Relative frequency of plot picks compared to other plots in the lineup plotted against the $p$-value (on log$_{10}$ scale) of each plot for all of the 60 individual lineups for experiment 1. Red indicates the plot with the lowest $p$-value, and blue indicates the actual data plot, when it is different from that with the lowest $p$-value. The columns of both top and bottom panel are ordered according to effect size, rows within panels show replicates for each parameter combination.}
       \label{fig:P-val_log}
\end{figure*}

\begin{figure*}[hbtp]
   \centering
       \includegraphics[width=0.95\textwidth]{p_val_log_counts2-a.pdf}       
       \includegraphics[width=0.95\textwidth]{p_val_log_counts2-b.pdf}
       \caption{Relative frequency of plot picks compared to other plots in the lineup plotted against the $p$-value (on log$_{10}$ scale) of each plot for all of the 70 individual lineups for experiment 2. Red indicates the plot with the lowest $p$-value, and blue indicates the actual data plot, when it is different from that with the lowest $p$-value. 
Columns are ordered according to effect size, with rows showing replicates of the same parameter combination on top of each other.
       Empty cells indicate combination of parameters with fewer than five replicates. 
       Highest counts tend to be the plot in the lineup having the lowest $p$-value, even more so than for experiment 1.
       }
       \label{fig:P-val_log2}
\end{figure*}

@@

We have seen that the subjects tend to pick the plot in the lineup that has the lowest $p$-value (Figures \ref{fig:P-val_log}, \ref{fig:P-val_log2}). What we are also interested in  is how this pick is affected by the distribution  of $p$-values of other plots in the lineup, particularly the $p$-value of the null plot with the strongest structure. If there is a null plot with a small $p$-value, or one close to that of the actual data plot, we would expect that subjects have a harder time detecting the actual data plot. Figure \ref{fig:pval_difference} investigates this. The difference between the $p$-value of the actual data is compared with the lowest from the null plots. This is plotted horizontally, and the proportion correct is plotted vertically. Negative values indicate lineups where the actual data plot had a smaller $p$-value than the minimum of the null plots. In experiment 1 (boxplots) there were a lot of lineups where the actual data plot had the smallest $p$-value, but only just. This caused quite some confusion for subjects, as seen because the variability in the proportion correct is huge for these lineups. Similarly large variability in correctness can be seen in the results of experiment 2 (scatterplots) except that the greater range of of differences in $p$-values shows the strength of subject's ability to pick the plot with most structure. Figures \ref{fig:P-val_log}, \ref{fig:P-val_log2} shed some more light on this story: when there is a big difference between the $p$-values (eg experiment 1, $\beta>7$) the subjects as one force chose the same plot. When there is less difference the distribution of counts is much more evenly spread between plots (eg experiment 1, $\beta=1$).

>>>>

\begin {abstract}  
Statistical graphics play a crucial role in exploratory data analysis, model checking and diagnosis. A new method, called the lineup protocol, enables determining the statistical significance of visual findings, which bridges a big gulf between exploratory and inferential statistics.  In this paper inferential methods for statistical graphics is developed further by refining the terminology of visual inference, and framing the lineup protocol in a context that allows direct comparison with conventional tests, in scenarios when a conventional test exists. This framework is used to compare the performance of the lineup protocol against conventional statistical testing in the scenario of fitting linear models. A human subjects experiment is conducted using simulated data to provide controlled conditions. Results suggest that the lineup protocol provides comparable results to the conventional tests, and expectedly out-performs them when data is contaminated, a scenario where assumptions required for performing a conventional test are violated, and there may be some super-individuals who yield better power than the conventional test even in the most difficult tasks.

@@

\begin{proof} The proof and details of the lemma is attached in the supplementary documents. 
\end{proof}

%\begin{proof}
%%We will now make use of properties of the data going into the lineup and assume that the properties are reflected in the lineup display:
%
% By definition  
%
%$$p_D=Pr\left(|t| \ge t_{obs} \mid H_0\right)=1-F_{|t|}(t_{obs}) \ \ \Rightarrow \ \  |t_{obs}|=F_{|t|}^{-1}(1-p_D)$$
%
%%Now suppose $F_{|t|}(.;\delta)$ denotes the distribution function of an absolute value of $t$, the conventional test statistic.
%%, with non-centrality parameter $\delta$. 
%\noindent Then the distribution function of the $p$-value, $p_D$, under $H_0$, is uniform, since:  
%\begin{eqnarray}\label{dist_p}
%F_{p_D}(p) &=& Pr(p_D \le p)=1-Pr(1-p_D \le 1-p) \nonumber \\
%  &=& 1-Pr\left(F_{|t|}^{-1}(1-p_D) \le F_{|t|}^{-1}(1-p) \right) \nonumber \\
%  &=& 1-Pr\left(|t_{obs}| \le F_{|t|}^{-1}(1-p)\right) \nonumber \\
%  &=&%\left\{ \begin{array}{ll}
%          1-F_{|t|}\left( F_{|t|}^{-1}(1-p)\right)=p \mbox{ ; under $H_0$} 
% %         1-F_{|t|}( F_{|t|}^{-1}(1-p); \delta) &\mbox{ ; under $H_1$} 
%%       \end{array} \right.     
%\end{eqnarray}
%
%%Thus the density of $p_D$ is Uniform(0,1)  under $H_0$. As noted by \cite{Ruppert:2007}, under $H_1$ the density of $p_D$ $$f_{p_D}(p_D; \delta)= \frac{f_{|t|}(F_{|t|}^{-1}(1-p_D);\delta)}{f_{|t|}(F_{|t|}^{-1}(1-p_D))}$$ derived from equation \eqref{dist_p}  is a right skewed distribution.  
%
%
%
%Let $p_{0,i}$, $i=1, ..., m-1$ denote the  $p$-values associated with data corresponding to the $m-1$ null plots. Since this data is generated consistently with the null hypothesis,  the $p$-values are independent and  follow a standard Uniform distribution, $p_{i,0} \sim U[0,1], i= 1, ..., m-1$. The minimum $p_0 = \min_{1 \le i \le m-1}  \ p_{0,i}$ then follows a Beta distribution with shape parameters 1 and $m-1$, and corresponding distribution function 
%\[
%F_{p_0} (x) =  1- (1- x) ^{m-1}  \text{ for } x \in [0,1].
%\]
%
%\noindent Thus
%\begin{eqnarray*}
%P(p_D < p_{0}) &=& 1 - P(p_{0} \le p_D) = 1- \int_0^1  P(p_{0} \le p_D \mid p_D=t) f_{p_D}(t) dt   \\
%&=& 1 - \int_0^1 F_{p_{0}}(t) f_{p_D}(t) dt = 1 - \int_0^1 f_{p_D}(t) dt + \int_0^1 (1-t)^{m-1} f_{p_D}(t) dt  \\
%&=&  E\left[ (1 - p_D)^{m-1}\right].
%\end{eqnarray*}
%
%
%%\noindent This proves the statement above. 
%%We further see:
%%\[
%%E\left[ (1 - p_D)^{m-1}\right] =  \sum_{k=0}^{m-1} {m-1 \choose k} (-1)^k E[p_D^k] = 1-(m-1)E[p_D] + O(E[p_D^2]).
%%\]
%
%
%% If the density of $p$-value is very right skewed, the expectation term would be large. The distribution function would be highly right skewed when there would be strong signal in the data plot.Thats where the distribution function of $p$-value under alternative comes in. Should we add that in Equation \ref{dist_p}? )},
%%conversely, as $m$, the number of choices given in the lineup, goes up the probability to pick the data plot  goes down. 
%%}
%%\blue{This shows that the lineup size $m$ has an effect on the power of the visual test: it is the $m-1$th moment of the distribution of $1-p_D$. 
%
%%
%%
%%
%%Let us think of a lineup as a head-to-head comparison of the test statistic and $m-1$ null plots. 
%%%Let the data plot have a $p$-value of $p_D$, and the null plots $p$ values of $p_{0, i}$ with $1 \le i \le m-1$.
%%We know that for each comparison the probability that the data, on which the plot is based, has the smaller $p$ value is 
%%\begin{eqnarray*}
%%P(p_D < p_{0,i}) &=& 1 - P(p_{i,0} \le p_D) = 1- \int_0^1  P(p_{i,0} \le p_D \mid p_D=t) f_{p_D}(t) dt =  \\
%%&=& 1 - \int_0^1 F_{p_{0,i}}(t) f_{p_D}(t) dt = 1 - \int_0^1 t f_{p_D}(t) dt = 1 - E[p_D],
%%\end{eqnarray*}
%%which, in particular, is independent of $p_{0,i}$ for all $i$.
%%
%%Let us make the assumption that an observer is able to identify  the chart corresponding to the data with the smallest $p$-value. Further we will assume that all observers  have the same ability in identifying the data plot.
%%
%%With that, we define $Z$  as the number of null plots in a lineup, for which the $p$-value $p_{0,i}$ is smaller than $p_D$.  
%%
%%Then $Z \sim B_{m-1, E[p_D]}$, and the probability that an observer will pick the data plot in a given lineup is 
%%\[
%%P(Z=0) = \left(1 - E[p_D] \right)^{m-1} = P(p_D \le p_0), \ \ \ \text{ where } p_0 = \min_{1 \le i \le m-1}  \ p_{0,i}.
%%\]
%\end{proof}

@@

Figure \ref{fig:P-val_log2} shows the selection counts and $p$-values of all plots for each lineup in experiments 1 and 2, respectively. In each of these plots, the log of $p$-value for each plot in the lineup is plotted horizontally and the count of the number of subjects who selected this plot is displayed vertically. The dot represents this point, and the lines connect vertically with 0 count. Columns are ordered according to effect size as defined in section 6.2; rows show replicates for the same combination of parameters.

@@

%\begin{figure*}[hbtp]
%   \centering
%      \includegraphics[width=0.8\textwidth]{p_val_log_counts-a.pdf}
%       \includegraphics[width=0.8\textwidth]{p_val_log_counts-b.pdf}
%       \caption{Relative frequency of plot picks compared to other plots in the lineup plotted against the $p$-value (on log$_{10}$ scale) of each plot for all of the 60 individual lineups for experiment 1. Red indicates the plot with the lowest $p$-value, and blue indicates the actual data plot, when it is different from that with the lowest $p$-value. The columns of both top and bottom panel are ordered according to effect size, rows within panels show replicates for each parameter combination.}
%       \label{fig:P-val_log}
%\end{figure*}

\begin{figure*}[hbtp]
   \centering
      \includegraphics[width=0.8\textwidth]{p_val_log_counts-a.pdf}
       \includegraphics[width=0.8\textwidth]{p_val_log_counts-b.pdf}   
       \includegraphics[width=0.8\textwidth]{p_val_log_counts2-a.pdf}       
       \includegraphics[width=0.8\textwidth]{p_val_log_counts2-b.pdf}
       \caption{ Relative frequency of plot picks compared to other plots in the lineup plotted against the $p$-value (on log$_{10}$ scale) of each plot for all individual lineups of both experiment 1 and 2. Red indicates the plot with the lowest $p$-value, and blue indicates the actual data plot, when it is different from that with the lowest $p$-value. 
Columns are ordered according to effect size, with rows showing replicates of the same parameter combination on top of each other. Empty cells indicate combination of parameters with fewer than five replicates. Highest counts tend to be the plot in the lineup having the lowest $p$-value, even more so than for experiment 1.
       }
       \label{fig:P-val_log2}
\end{figure*}

@@

We have seen that the subjects tend to pick the plot in the lineup that has the lowest $p$-value (Figure \ref{fig:P-val_log2}). What we are also interested in  is how this pick is affected by the distribution  of $p$-values of other plots in the lineup, particularly the $p$-value of the null plot with the strongest structure. If there is a null plot with a small $p$-value, or one close to that of the actual data plot, we would expect that subjects have a harder time detecting the actual data plot. Figure \ref{fig:pval_difference} investigates this. The difference between the $p$-value of the actual data is compared with the lowest from the null plots. This is plotted horizontally, and the proportion correct is plotted vertically. Negative values indicate lineups where the actual data plot had a smaller $p$-value than the minimum of the null plots. In experiment 1 (boxplots) there were a lot of lineups where the actual data plot had the smallest $p$-value, but only just. This caused quite some confusion for subjects, as seen because the variability in the proportion correct is huge for these lineups. Similarly large variability in correctness can be seen in the results of experiment 2 (scatterplots) except that the greater range of of differences in $p$-values shows the strength of subject's ability to pick the plot with most structure. Figure \ref{fig:P-val_log2} shed some more light on this story: when there is a big difference between the $p$-values (eg experiment 1, $\beta>7$) the subjects as one force chose the same plot. When there is less difference the distribution of counts is much more evenly spread between plots (eg experiment 1, $\beta=1$).
